---
title: "Gene Expression IMRaD Report"
format: html
editor: source
author: "Vivienne Niejalke"
date: "16 May 2023"
execute: 
  echo: false
---

```{r}
pacman::p_load(lme4,ggplot2)

#read in data
gene<-read.csv(here::here("data/2023_03_21-gene_expression.csv"))
```

## Introduction

The dataset is gene expression on growth factor based on treatments. The key parameters are treatment, concentration and cell type. The gene expression is evaluated with growth factor. The aim of this analysis is to make a predictive model for gene expression.

## Methods

I did the work in R using the lme4 package to make a linear mixed effect random effects model. Consider doing an AIC and commenting on why there is no cell type term. Show summary stats and that cell type isn't that important. Label plot better and include plot from previous assignment. Different shapes for treatment and cell type (distinguish in the plot)

## Results
Here are some nice plots.

```{r}
model_lme4<-lmer(growth_factor~(concentration|gene_expression)+ concentration+treatment,data=gene)

#plot predictive model
ggplot(gene,aes(x=concentration,y=growth_factor,color=gene_expression))+
  geom_point()+
  geom_line(aes(y=predict(model_lme4)))+
  labs(
    x="\u00b5g/ml",
    y="Growth Factor (units)"
  ) + ggtitle("Predictive Model for Gene Expression")

```

## Discussion
The treatment is an important factor for gene expression.