---
title: Gene Expression IMRaD Report
format: pdf
editor: source
author: Vivienne Niejalke
date: today
execute: 
  echo: false
bibliography: gene.bib
---

```{r}
pacman::p_load(lme4,ggplot2,lmerTest,
               RColorBrewer,ghibli)

#read in data
gene<-read.csv(here::here("data/2023_03_21-gene_expression.csv"))
```

## Introduction

The dataset contains 87 observations of an experiment testing the effects of a new treatment on gene expression. This data was provided by Dr Karl Berator, a research assistant at the Institute of -omics. The key variables were:

- treatment: the new treatment was compared with a saline placebo, 
- concentration: an integer between 0 and 10, measured in $\mu g/ml$, 
- gene line: there were eight specific genes, and
- cell type: each gene line was classified as Wild-type or Cell-type 101. 

The aim was to investigate the effect of the growth factor on gene expression. This was evaluated by designing a predictive model for gene expression.

## Methods

The data was analysed using the packages `lme4` [@lme4-2023], `lmerTest` [@lmerTest-2023] and `ggplot2` [@ggplot2-2023] in R [@r-2023] in RStudio[^1].

[^1]: https://www.rstudio.com

The data was manually preprocessed to combine the data from for each gene line into one dataset.

Firstly to better understand the data, an exploratory data analysis (EDA) was conducted to visualize the data using box plots and plots.

Then, a linear random mixed effects model with a random intercept term was fitted to the data, with concentration, treatment and cell type as predictors and gene line as the random intercept term. The joint interaction term is significant due to the p-value less than 0.05.

```{r}
model_1<-lmer(growth_factor~(1|gene_expression)+ concentration*treatment*cell_type,data=gene)


```

A linear random mixed effects model with a random slope was also fitted, with concentration, treatment and cell type as predictors and concentration per gene line as the random slope term.

```{r}
model_2<-lmer(growth_factor~(concentration|gene_expression)+ concentration*treatment*cell_type,data=gene)

#interaction terms not significant - becomes model 1 - need to remember backstep
model_3<-lmer(growth_factor~(concentration|gene_expression)+ concentration+treatment,data=gene)

```

The factors of the two models were assessed using ranova for random effects and anova for fixed effects. The models were compared using the AIC (Akaike Information Criterion) to select the best model.

## Results
Figure 1 of the box plots show the gene expression of different gene lines and cell types. It shows that the treatment has a greater gene expression, however greater variability in the results.

```{r, fig.cap="Box Plot of the Effect of Treatment and Cell Line on Gene Expression"}
ggplot(gene,aes(x=treatment,y=growth_factor,fill=treatment,color=gene_expression))+geom_boxplot() +
  facet_wrap(~cell_type)+scale_fill_ghibli_d(name="MononokeMedium",direction=-1)+scale_color_brewer(name="Gene",palette = "Dark2") +
  labs(x="Treatment",y="Gene Expression")+
  ggtitle("Effect of Treatment and Cell Line on Gene Expression")

```

The 'ranova' and 'anova' tests indicated that the models are suitable and all factors are relevent, especially due to the significance of the high order interaction terms. The table below indicates that 'model_2', the random slope mixed effects model with interaction terms is the better predictive model.

```{r}
#analyse it using AIC
AIC(model_1,model_2,model_3)

```

The selected linear mixed effects model with a random slope is plotted below as the chosen predictive model.This shows that the gene expression is not only greater for genes with the treatment, but also increases as the concentration increases.

```{r, fig.cap="Linear Mixed Effects Random Model"}

#plot predictive model
ggplot(gene,aes(x=concentration,y=growth_factor,color=gene_expression))+
  geom_point(aes(shape=treatment))+
  geom_line(aes(y=predict(model_2)))+
  labs(
    x="\u00b5g/ml",
    y="Gene Expression"
  ) + ggtitle("Predictive Model for Gene Expression")+
  scale_color_brewer(name="Gene",palette = "Dark2")

```


## Discussion
From the provided plots, it is illustrated that the treatment increased the growth factor of the gene expression for both cell types. Concentration, treatment and cell type are considered significant factors for the predictive model.

The AIC results indicate that a linear mixed effects model with a random slope is better than a linear mixed effects model with only a random intercept. This means that the correlated concentration and gene line random effect is important to consider.

## References
